---
title: "Data filtering"
author: "Ricardo H. Ramirez Gonzalz"
date: "14/12/2023"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Installing Dependencies in R

For this tutorial you need a few R libraries. Bioconductor has packages for bioinfmatics. The following command will install the libraries used in the  tutorial. 

```{r warning=FALSE, error=FALSE, message=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.18")
BiocManager::install(c("rtracklayer", "ggbio", "knitr", "ggplot2", "tidyr", "dplyr","VariantAnnotation"))

```

The documentation of the packages used is in: 

* https://bioconductor.org/packages/release/bioc/vignettes/VariantAnnotation/inst/doc/VariantAnnotation.pdf
* https://dplyr.tidyverse.org
* https://www.tidyverse.org

# Reading the VCF file

The VCF file is a file format that contains SNPs in a database. We will be working with a selection of SNPs from the following paper:

```
Elizabeth S. A. Sollars, et al (2016) Genome sequence and genetic diversity of European ash trees. Nature 541: 212–216 doi:10.1038/nature20786
```


## Loading the data

For this analysis we will need several R libraries to do the analysis, filtering and plot. 

```{r warning=FALSE, error=FALSE, message=FALSE}
library(rtracklayer)
library(knitr)
library(ggplot2)
library(ggbio)
library(tidyr)
library(dplyr)
library(VariantAnnotation)
vcf <- readVcf(file = "extractedContig1.6.23.vcf.gz")
```



# VCF header

You can use the following command to see all the data that is stored in the VCF

```{r, echo = TRUE}
head(vcf)
```

# Looking at the qualities of the

To get the a ```GRanges``` object with the common data the function ```rowRanges``` can be used. 
This let us query questions like the distribution of the quality of the SNPs.

```{r, echo=TRUE}
variations <- rowRanges(vcf)
head(variations, 3)
```

# Summary of qualities

```{r, echo=TRUE}
summary(variations$QUAL)
```

## Visalising the distribution

To get an idea of the overall quality of the data, a histogram is useful. 
The SNPs with lower quality tend to be noisy. 
For this project, we validated some random SNPs using KASP across different qualitues, from 50 to 500 and we concluded that 300 was a good spot. 
This histogram helped us to decided the thresholds to test, as most of the tutorials online deem QUAL>50 as good quality. 

```{r}
ggplot(variations, aes(QUAL)) + geom_histogram(binwidth = 50) + theme_bw() +
   coord_cartesian(xlim=c(0,20000)) + geom_vline(xintercept=300, color="red")
```


# Depth for individual samples

We can look at the depth of coverage of various samples. This can be useful to find if a particular sample has low quality. 

```{r}
DP <-geno(vcf)$DP 
summary_dp <- DP[is.na(DP)] <- 0
long_dp <- gather(data.frame(DP),value = "DP", key="Sample")
long_dp %>%
  group_by(Sample) %>%
  summarise(
    n = n(),
    mean_DP = mean(DP, na.rm = TRUE),
    std_DP  = sd(DP, na.rm = TRUE)
  ) -> summary_dp

```

```{r, echo=FALSE}
kable(summary_dp)
```

* Can you make a summary per contig?
* Can you make a summary per contig and per sample?

# Distribution across sample

You can focus on the distribution of a few samples.

```{r }
long_dp %>% filter( Sample %in% c("Fex1", "Fex2")) %>% 
ggplot() +  
geom_density(aes(x=DP, colour=Sample, fill=Sample), alpha=0.5) + 
coord_cartesian(xlim = c(0,100))
```

# Displaying all the samples

For all the samples, you can show just a boxplot,at the cost of having less resultion of the shape of the distribution


```{r }
long_dp %>% 
ggplot() +  
geom_boxplot(aes(y=DP, x=Sample), alpha=0.5) + 
coord_cartesian(ylim = c(0,25)) + 
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

# Questions for you 

- How many ```seqnames``` are present?
- Plot the count of genotypes on each sequence.
- Can you calculate the allele frequency ($\frac{R0}{DP}$) for each sample? Plot 2 or 3 samples.  
- Plot the distribution of several samples together (Team1: Sample Fex1-Fex5, Team 2: Fex6-Fex10, etc)
- Are there differences in in DP across different scaffolds? Try to figure out how to extract the data for a single reference. 
  - Maybe directly from the ```VCF``` object, or exporting the DP matrix and filter by the ```rowname```. 
- Do you think the samples are homozygous, heterozygous or a mix of both (you can calculate the distribution of allele frequencies or look at the GT column). 
- To compare all the sample, overlapping all the distributions looks messy. Plot the ```DP``` for all the samples. Try ```geom_boxplot``` and ```facet_grid```. Which one you find clearer? Is there any sample that has a significant lower coverage?

# How many ```seqnames``` are present?

```{r}
seqnames(vcf)
```

# Plot the count of genotypes on each sequence.

We need to produce a dataframe with the contig names and the ```GT``` column

```{r}
GT <-geno(vcf)$GT
GT <- data.frame(GT)
GT$seqnames <- data.frame(seqnames(variations))[,1]
GT %>% gather( -seqnames ,key = "Sample", value = "GT") -> long_gt

long_gt %>%
  group_by(Sample, GT, seqnames) %>%
  summarise(
    n = n(),
    GT = unique(GT, na.rm = TRUE)
  )
```

#  Plot the count of genotypes on each sequence.

Now we can plot from the ```GT``` data frame and plot the counts

```{r}
 long_gt %>% 
filter( Sample %in% c("Fex1", "Fex2", "Fex3", "Fex4", "Fex5")) %>% 
ggplot( aes(GT, fill=GT)) + geom_bar( ) + facet_grid(seqnames~Sample)
```

# Can you calculate the allele frequency ($\frac{RO}{DP}$) for each sample? Plot 2 or 3 samples.  

```{r}
DP <-geno(vcf)$DP 
RO <-geno(vcf)$RO 
ro_freq <- data.frame(RO/DP)

ggplot(ro_freq, aes(Fex1)) + geom_histogram()

```

# Plot the distribution of several samples together (Team1: Sample Fex1-Fex5, Team 2: Fex6-Fex10, etc)

```{r}
long_ro_freq <- gather(ro_freq ,value = "RO_freq", key="Sample")
long_ro_freq %>% filter( Sample %in% c("Fex1", "Fex3","Fex4", "Fex2","Fex5","Fex6")) %>% 
ggplot() +  
geom_histogram(aes(x=RO_freq), alpha=0.5) + facet_wrap(Sample ~ .)
```

# Are there differences in in DP across different scaffolds? Try to figure out how to extract the data for a single reference. 

Hint: directly from the ```VCF``` object, or exporting the DP matrix and filter by the ```rowname```. 

```{r}
DP <-geno(vcf)$DP
DP <- data.frame(DP)
DP$seqnames <- data.frame(seqnames(variations))[,1]
DP %>% gather(  -seqnames ,value = "DP", key="Sample") -> long_dp
```

The coverage is very similar across sequences

```{r}
long_dp %>%
  group_by(seqnames) %>%
  summarise(
    n = n(),
    DP = mean(DP, na.rm = TRUE)
  )
```

Most of the variation comes from the sample

```{r}
long_dp %>%
  group_by(Sample) %>%
  summarise(
    n = n(),
    DP = mean(DP, na.rm = TRUE)
  )
```


# To think about...

- Do you think the samples are homozygous, heterozygous or a mix of both (you can calculate the distribution of allele frequencies or look at the GT column). 

The samples look homozygous, with small regions with heterozygous calls.

- To compare all the sample, overlapping all the distributions looks messy. Plot the ```DP``` for all the samples. Try ```geom_boxplot``` and ```facet_grid```. Which one you find clearer? Is there any sample that has a significant lower coverage?
This is a personal choice. The plots above reflect it. 

# Filtering. 

Based on the discussion before, using ```bcftools``` filger out for files with quality over 300. 
Read the documentation in: https://samtools.github.io/bcftools/bcftools.html#expressions to find more expression. 

First simplify the VCF file to only include the GT,AO and RO files. 

```sh
bcftools annotate -x "INFO,^FORMAT/GT,FORMAT/AO,Format/RO" extractedContig1.6.23.vcf.gz | bgzip -c > extractedContig1.6.23.anno.vcf.gz
```
beagle requires that even missing values correspond to the ploidy, so we need to change ```.``` in the genotype to ```./.```

```sh
gunzip -c extractedContig1.6.23.anno.vcf.gz | sed -e 's/    \.:/    .\/.:/g' | bgzip -c >  extractedContig1.6.23.anno.fixed.vcf.gz
```

Now we can run the filtering of the formatted file
```sh
bcftools filter -e 'QUAL<300' extractedContig1.6.23.anno.fixed.vcf.gz  -O z -o extractedContig1.6.23.anno.fixed.QUAL300.vcf.gz
```
- Can you design a filter to remove only with heterozygous calls on  line ```Fex10```? 
Hint: look for ```FORMAT/DP``` in the documentation above. 

- How does the number of heterozygous calls are affected on each line after filtering?
Hint: You  need to look at the GT tutorial above. 


# Phasing. 

We will use beagle to do the phasing and the imputing: 
http://faculty.washington.edu/browning/beagle/beagle.html

The documentation is here:
http://faculty.washington.edu/browning/beagle/beagle_5.1_08Nov19.pdf

```sh
java -jar beagle.21Sep19.ec3.jar gt=extractedContig1.6.23.anno.fixed.QUAL300.vcf.gz  out=out.gt
```

Now, you can load ```out.gt.vcf.gz``` to see the results

- Do you see a change in the number of heterozygous samples?

