---
title: "Lecture 11: Comparison and prediction of protein structure"
author:
- affiliation: Fundación ARAID & Estación Experimental de Aula Dei-CSIC, Zaragoza
    España <http://www.eead.csic.es/compbio>
  name: Bruno Contreras Moreira
date: '`r format(Sys.Date(),"%d/%b/%Y")`'
output:
  html_document:
    fig_caption: yes
    highlight: zenburn
    number_sections: no
    theme: cerulean
    toc: yes
    toc_float: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
geometry: margin=1cm
fontsize: 11pt
---

```{r knitr setup, include=FALSE,  eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
knitr::opts_chunk$set(eval=TRUE, cache=FALSE, message=FALSE, warning=FALSE, 
                      comment = "", results="markup")
```

## Exercise 11.1

Please browse the [Protein Data Bank](https://www.rcsb.org) and:

  * Choose a protein and explore its structure with the "3D View" tab. This will be the **query protein**. I suggest 9ANT just in case.
  
  * Download the corresponding file in PDB format.
  
  * Download the corresponding sequence file in FASTA format.
  
  * Install R package [Rpdb](https://cran.r-project.org/web/packages/Rpdb/index.html), load the PDB file and visualize it.
  
  * Install [PyMOL](https://pymol.org) and visualize the PDB.

## Exercise 11.2

Visit the [HHpred](https://toolkit.tuebingen.mpg.de) web server and launch a structure prediction job with the FASTA sequence obtained above against the nr70 PDB collection:

  * Add to your report the secondary structure prediction.

  * Save the alignments and scores of the top 3 PDB structures.
  
  * Visit the PDB web site and download the PDB files of these top structures/templates.

## Exercise 11.3

Download the FORTRAN source code of TM-align at [https://zhanglab.ccmb.med.umich.edu/TM-align](https://zhanglab.ccmb.med.umich.edu/TM-align):

  * Compile *TM-align* with 
  
```{r fortran, eval=FALSE}
gfortran -static -O3 -ffast-math -lm -o TMalign TMalign.f
```

  * Compute the TM-score and RMSD of the **query structure** when compared to all top structures with *TM-align* and build a table with them.


  * Save a PyMOL picture of each of three structure superpositions.