---
title: "Analysis of transcriptomes. WGCNA tool for D dataset"
author: "Rubén Sancho"
date: "2025-12-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = TRUE)
```

```{r install_WGCNA_v.1.73}

# https://cran.r-project.org/web/packages/WGCNA/index.html
# https://cran.r-project.org/web/packages/WGCNA/WGCNA.pdf

#if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("GO.db")

#install.packages("WGCNA")

```

```{r WGCNA}

library(WGCNA)

#help(package = 'WGCNA')

```


```{r dataset}

setwd(".")

getwd()

# Drought (D)

D_dataset = read.csv("TPM_counts_Drought_D_dataset.csv")
dim(D_dataset)
names(D_dataset)

# Transpose rows/columns in D dataset

datExprD = as.data.frame(t(D_dataset[, -c(1)]));
names(datExprD) = D_dataset$target_id;
rownames(datExprD) = names(D_dataset)[-c(1)];

```
# Iterative filtering of samples and genes with too many missing entries

```{r goodgenes_D}

# Run all chunk

# remove genes with too many missing samples in D

gsgD = goodSamplesGenes(datExprD, verbose = 3);

summary(gsgD)

gsgD$allOK # if TRUE, all genes are ok

if (!gsgD$allOK)
{
  # Optionally, print the gene and sample names that were removed:
  if (sum(!gsgD$goodGenes)>0) 
     printFlush(paste("Removing genes:", paste(names(datExprD)[!gsgD$goodGenes], collapse = ", ")));
  if (sum(!gsgD$goodSamples)>0) 
     printFlush(paste("Removing samples:", paste(rownames(datExprD)[!gsgD$goodSamples], collapse = ", ")));
  datExprD = datExprD[gsgD$goodSamples, gsgD$goodGenes]
  dim(datExprD)
}
```

## Identifying outlier samples
## You can identify outlier samples by using hierarchical clustering.

```{r D_outliers}

# Run all chunk

# cluster and plot the samples (genes that will come later) to see if there are any obvious outliers

# D dataset outliers detection

sampleTreeD = hclust(dist(datExprD), method = "average")
sizeGrWindow(12,9)
#png("dendrogram_D.png", width = 1600, height = 1200, units = "px", pointsize = 32, bg = "white");
par(cex = 0.6);
par(mar = c(0,4,2,0))
plot(sampleTreeD, main = "D Sample clustering to detect outliers", sub="", xlab="", cex.lab = 1.5, 
     cex.axis = 1.5, cex.main = 2)
outliersD=300000
abline(h=outliersD, col = "red"); # to mark/cut  outliers
#dev.off()

```

# Save filtered dataset

```{r filter_dataset}

## keep only desired samples in D dataset

clustD = cutreeStatic(sampleTreeD, cutHeight=outliersD, minSize = 20)
table(clustD)
keepSamplesD = (clustD==1)
datExprD = datExprD[keepSamplesD, ]
nGenesD = ncol(datExprD)
nSamplesD = nrow(datExprD)

## keep filtering dataset

save(datExprD, file = "datExpr_D.RData")

```

# Soft-thresholding powers selection for network construction

```{r D_power}

# Choose a set of soft-thresholding powers

powers = c(c(1:13), seq(from = 14, to=20, by=2))

# Analysis of scale free topology for soft-thresholding

# Unsigned: Both strong positive and strong negative correlations 
#           are considered connections
# Signed: Only positive correlations contribute to network connectivity.
#         Negative correlations reduce similarity toward zero, but correlations 
#         near zero still retain some intermediate similarity.
# Signed hybrid: Positive correlations are treated the same as 
#                in signed networks, but negative correlations are 
#               truncated to zero (completely ignored).

sftD = pickSoftThreshold(datExprD,
                         powerVector = powers,
                         networkType = "unsigned", # default
                         verbose = 5)
```

```{r Plot_D_power}

# Plot the results:
sizeGrWindow(9, 5)
par(mfrow = c(1,2));
cex1 = 0.9;

# Scale-free topology fit index as a function of the soft-thresholding power D

#png("power_D_pearson_unsigned.png", width = 1600, height = 1200, units = "px", 
#    pointsize = 32, bg = "white");
plot(sftD$fitIndices[,1], -sign(sftD$fitIndices[,3])*sftD$fitIndices[,2],
xlab="Soft Threshold (power)",ylab="Scale Free Topology Model Fit,signed R^2",
type="n",
main = paste("Scale independence (D)"));
text(sftD$fitIndices[,1], -sign(sftD$fitIndices[,3])*sftD$fitIndices[,2],
labels=powers,cex=cex1,col="red");

# this line corresponds to using an R^2 cut-off of h
abline(h=0.85,col="red")
#dev.off()

# Scale-free topology connectivity
#png("connectivity_D_pearson_unsigned.png", width = 1600, height = 1200,
#    units = "px", pointsize = 32, bg = "white");

plot(sftD$fitIndices[,1], sftD$fitIndices[,5],
     xlab="Soft Threshold (power)",ylab="Mean Connectivity", type="n",
     main = paste("Mean connectivity (D)"))
text(sftD$fitIndices[,1], sftD$fitIndices[,5], labels=powers, cex=cex1, 
     col="red")

#dev.off()

```
## Optimal D power = 5

# Unsigned net for D dataset

```{r net_D}

# load dataset

load(file = "datExpr_D.RData");

genes = colnames(datExprD)

# Calling the Adjacency Function

adjacency_D <- adjacency(datExprD, power = 5)

# Topological Overlap Matrix

TOM_D <- TOMsimilarity(adjacency_D)   ## This step takes a long time (5 min) ##

# convert this matrix into a dissimilarity matrix 

TOM_diss_D <- 1-TOM_D

# Hierarchical Clustering Analysis

# Creating the dendrogram 

geneTree_D <- hclust(as.dist(TOM_diss_D), method = "average")

# Plotting the dendrogram

sizeGrWindow(12,9)
plot(geneTree_D, xlab="", sub="", 
     main = "DROUGHT: Gene clustering on TOM-based dissimilarity", 
labels = FALSE, hang = 0.04)

# Identification of modules D

modules_D <- cutreeDynamic(dendro = geneTree_D, distM = TOM_diss_D, 
                           deepSplit = 2, pamRespectsDendro = FALSE, 
                           minClusterSize = 30)


table(modules_D) #returns a table of the counts of factor levels in an object. 
            # In this case how many genes are assigned to each created module.


modulecolors_D <- labels2colors(modules_D) # assigns each module number a color

table(modulecolors_D) # returns the counts for each color
                      # (aka the number of genes within each module)


# Plots the gene dendrogram with the module colors

plotDendroAndColors(geneTree_D, modulecolors_D,"Module D",
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05,
main = "Gene dendrogram and module colors for D network")


# Module Eigengene Identification
## A ME (Module Eigengene) is the standardized gene expression profile for 
## a given module.

## An eigengene is the gene whose expression is representative of the the 
## majority of genes expressed within a module.

MElist_D <- moduleEigengenes(datExprD, colors = modulecolors_D) 
MEs_D <- MElist_D$eigengenes 

head(MEs_D)
```

```{r merging}

# Module Merging

ME_diss_D = 1-cor(MElist_D$eigengenes, 
                         use="complete") # Calculate eigengene dissimilarity

METree_D = hclust(as.dist(ME_diss_D), 
                  method = "average") # Clustering eigengenes 
par(mar = c(0,4,2,0)) # setting margin sizes
par(cex = 0.6); # scaling the graphic
plot(METree_D)
abline(h=.25, col = "red") # a height of .25 corresponds to correlation of .75


merge_D <- mergeCloseModules(datExprD, modulecolors_D, cutHeight = .25)


# The merged module colors, assigning one color to each module

mergedcolors_D = merge_D$colors

# Eigengenes of the new merged modules

mergedMEs_D = merge_D$newMEs

# Plot the merged dendrogram

plotDendroAndColors(geneTree_D, cbind(modulecolors_D, mergedcolors_D), 
c("Original Module D", "Merged Module D"),
dendroLabels = FALSE, hang = 0.03,
addGuide = TRUE, guideHang = 0.05,
main = "Dendrogram and modules for original and merged modules in D dataset")

# Rename to modulecolors

modulecolors_D = mergedcolors_D

# Construct numerical labels corresponding to the colors

colororder_D = c("grey", standardColors(50));
 
modulelabels_D = match(modulecolors_D, colororder_D)-1;
 
MEs_D = mergedMEs_D;

table(modulelabels_D) # Merged modules = 40; original modules = 60

table(modulecolors_D)

# Save module colors and labels for use in subsequent parts
 
save(MEs_D, modulelabels_D, modulecolors_D, geneTree_D, file = "net_D.RData")


```

# Export genes-modules

```{r export_modules_genes}

# Export lists of genes within each module of D

load("datExpr_D.RData")
load("net_D.RData")

# Extract gene names

geneNames <- colnames(datExprD)

# Create gene-module table

geneModuleMembership_D <- data.frame(
  Gene = geneNames,
  ModuleColor = modulecolors_D
)

head(geneModuleMembership_D)

# Prepare data frame and export to tsv file

genes_by_module_D <- split(geneModuleMembership_D$Gene, geneModuleMembership_D$ModuleColor)


genes_by_module_df_D <- data.frame(
  ModuleColor = rep(names(genes_by_module_D), times = sapply(genes_by_module_D, length)),
  Gene = unlist(genes_by_module_D, use.names = FALSE)
)

head(genes_by_module_df_D)

# Export to tsv

write.table(genes_by_module_df_D,
            file = "Genes_per_module_D.tsv",
            sep = "\t",        # sep tab
            quote = FALSE,
            row.names = FALSE,
            col.names = TRUE)

```

# Top hub gene per module

```{r Top_hub_gene}

# The single top hub gene per module:

chooseTopHubInEachModule(
   datExprD, 
   modulecolors_D, 
   omitColors = "grey", 
   power = 5, 
   type = "unsigned")

```

# Module-trait associations

```{r trait_dataset}

# Load expression data D

load("datExpr_D.RData")

# Load network data D

load("net_D.RData")

# Prepare trait D data set

traitdata <- read.csv("TRAITS_D.csv", header = TRUE, stringsAsFactors = FALSE)

head(traitdata)

# Keep samples and traits

allTraits <- traitdata[, c(4, 6:17) ] # keep samples and traits 
head(allTraits)
# variables/traits numeric

allTraits[, 2:ncol(allTraits)] <- lapply(allTraits[, 2:ncol(allTraits)], 
                                         function(x) as.numeric(as.character(x)))

# check classes

sapply(allTraits, class)

head(allTraits)

# Match the trait data to the expression data by the sample name

Samples <- rownames(datExprD)

traitRows <- match(Samples, allTraits$sample)

datTraits <- allTraits[traitRows, -1]

rownames(datTraits) <- allTraits[traitRows, 1]

# Eigengene gene significance: correlation of the trait with the identified 
# module eigengenes

# Define numbers of genes and samples

nGenes = ncol(datExprD)

nSamples = nrow(datExprD)

module.trait.correlation = cor(MEs_D, 
                               datTraits, 
                               use = "p") # Pearson correlation coefficient

# p-value associated with the correlation

module.trait.Pvalue = corPvalueStudent(module.trait.correlation, nSamples) 

# Will display correlations and their p-values

textMatrix = paste(signif(module.trait.correlation, 2), "\n(",
signif(module.trait.Pvalue, 1), ")", sep = "");

dim(textMatrix) = dim(module.trait.correlation)

pdf("heatmap_traits_modules_D.pdf", width = 8, height = 12) # save in pdf

par(mar = c(6, 8.5, 3, 1))

# Display the correlation values within a heatmap plot

labeledHeatmap(Matrix = module.trait.correlation,
                        xLabels = names(datTraits),
                        yLabels = names(MEs_D),
                        ySymbols = names(MEs_D),
                        colorLabels = FALSE,
                        colors = blueWhiteRed(50),
                        textMatrix = textMatrix,
                        setStdMargins = FALSE,
                        cex.text = 0.4,
                        zlim = c(-1,1),
                        main = paste("Module-trait relationships D"))

dev.off() # save plot in pdf

# The highest + correlation: MEgrey60 vs leafn (0.73 (p-value = 2e−06))
# The highest - correlation: MEpalevioletred3 vs rmr (−0.76 (p-value = 6e−07))

```